---
title: "proteoQC: an R package for proteomics data quality assessment"
author: "Bo Wen and Laurent Gatto"
date: "`r Sys.Date()`"
bibliography: proteoQC.bib
output: 
  prettydoc::html_pretty:
    toc: true
    theme: cayman
    highlight: github
vignette: >
  %\VignetteIndexEntry{00 proteoQC introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r style, echo=FALSE, results="asis", message=FALSE}
knitr::opts_chunk$set(tidy = FALSE,message = FALSE)
```



```{r echo=FALSE, results="hide"}
library("BiocStyle")
BiocStyle::markdown()
```




```{r echo=FALSE,warning=FALSE}
suppressPackageStartupMessages(library("proteoQC"))
suppressPackageStartupMessages(library("R.utils"))
```


# Introduction

The `proteoQC` package provides a integrated pipeline for mass
spectrometry-based proteomics quality control. It allows to generate a
dynamic report starting from a set of **mgf** or **mz[X]ML**
format peak list files, a protein database file and a description file of
the experimental design. It performs an MS/MS search against the protein
data base using the **X!Tandem** search engine [@Craig:2004] and the
`rTANDEM` package [@rTANDEM]. The results are then
summarised and compiled into an interactive html report using the
`Nozzle.R1` package [@Nozzle.R1,@Gehlenborg:2013].

# Example data

We are going to use parts a dataset from the ProteomeXchange
repository ([http://www.proteomexchange.org/](http://www.proteomexchange.org/)). We will use the
`rpx` package to accessed and downloaded the data.


```{r eval=TRUE,warning=FALSE,error=FALSE,cache=TRUE}
library("rpx")
px <- PXDataset("PXD000864")
px
```

There are a total of `r length(pxfiles(px))` files available from
the ProteomeXchange repository, including raw data files
(**raw**), result files (**-pride.xml.gz**), (compressed)
peak list files (**.mgf.gz**) and, the fasta database file
(**TTE2010.zip**) and one **README.txt** file.

```{r pxfiles, warning=FALSE,error=FALSE}
head(pxfiles(px))
tail(pxfiles(px))
```


The files, in particular the **mgf** files that will be used in
the rest of this document are named as follows **TTE-CC-B-FR-R**
where **CC** takes values 55 or 75 and stands for the bacteria
culture temperature in degree Celsius, **B** stands for the
biological replicate (only 1 here), **FR** represents the
fraction number from 01 to 12 and the leading **R** documents one
of three technical replicates. (See also
[http://www.ebi.ac.uk/pride/archive/projects/PXD000864](http://www.ebi.ac.uk/pride/archive/projects/PXD000864) for
details). Here, we will make use of a limited number of samples
below. First, we create a vector that stores the file names of
interest.

```{r eval=TRUE,warning=FALSE}
mgfs <- grep("mgf", pxfiles(px), value = TRUE)
mgfs <- grep("-0[5-6]-[1|2]", mgfs, value=TRUE)
mgfs
```

These files can be downloaded [^1] using the `pxget`, providing the
relevant data object (here `px`) and file names to be
downloaded (see `?pxget` for details). We also need to
uncompress (using `gunzip`) the files.



```{r eval=FALSE,cache=TRUE}
mgffiles <- pxget(px, mgfs)
library("R.utils")
mgffiles <- sapply(mgffiles, gunzip)
```

To reduce the file size of the demonstration data included for this
package, we have trimmed the peak lists to 1/10 of the original number
of spectra. All the details are provided in the vignette source.

```{r echo=FALSE, eval=FALSE}
## Generate the lightweight qc report, 
## trim the mgf files to 1/10 of their size.

trimMgf <- function(f, m = 1/10, overwrite = FALSE) {
    message("Reading ", f)
    x <- readLines(f)
    beg <- grep("BEGIN IONS", x)
    end <- grep("END IONS", x)
    n <- length(beg)
    message("Sub-setting to ", m)
    i <- sort(sample(n, floor(n * m)))
    k <- unlist(mapply(seq, from = beg[i], to = end[i]))
    if (overwrite) {
        unlink(f)
        message("Writing ", f)
        writeLines(x[k], con = f)
        return(f)
    } else {
        g <- sub(".mgf", "_small.mgf", f)
        message("Writing ", g)
        writeLines(x[k], con = g)
        return(g)
    }    
}

set.seed(1)
mgffiles <- sapply(mgffiles, trimMgf, overwrite = TRUE)
```

Similarly, below we download the database file and unzip it.

```{r eval=FALSE}
fas <- pxget(px, "TTE2010.zip")
fas <- unzip(fas)
fas
```


# Running `proteoQC`

```{r eval=FALSE, echo=FALSE}

## code to regenerate the design file
sample <- rep(c("55","75"),each=4)
techrep <- rep(1:2, 4)
biorep <- rep(1, length(mgffiles))
frac <- rep((rep(5:6, each = 2)), 2)
des <- data.frame(file = mgffiles,
                  sample = sample,
                  bioRep = biorep, techRep = techrep,
                  fraction = frac,
                  row.names = NULL)

write.table(des, sep = " ", row.names=FALSE,
            quote = FALSE,
            file = "../inst/extdata/PXD000864-design.txt")

```

## Preparing the QC

The first step in the `proteoQC` pipeline is the definition of a
design file, that provides the **mgf** file names,
**sample** numbers, biological **biocRep**) and technical
(**techRep**) replicates and **fraction** numbers in a
simple space-separated tabular format. We provide such a design file
for our `r length(mgfs)` files of interest.

```{r}
design <- system.file("extdata/PXD000864-design.txt", package = "proteoQC")
design
read.table(design, header = TRUE)
```

## Running the QC

We need to load the `proteoQC` package and call the
**msQCpipe** function, providing appropriate input parameters,
in particular the **design** file, the **fasta** protein
database, the **outdir** output directory that will contain the
final quality report and various other peptide spectrum matching
parameters that will be passed to the `rTANDEM` package. See
`?msQCpipe` for a more in-depth description of all its
arguments. Please note that if you take mz[X]ML format files as input, you must
make sure that you have installed the rTANDEM that the version is greater than
1.5.1.

```{r eval=FALSE, tidy=FALSE}
qcres <- msQCpipe(spectralist = design,
                  fasta = fas, 
                  outdir = "./qc",
                  miss  = 0,
                  enzyme = 1, varmod = 2, fixmod = 1,
                  tol = 10, itol = 0.6, cpu = 2,
                  mode = "identification")
```

The `msQCpipe` function will run each mgf input file
documented in the design file and search it against the fasta database
using the **tandem** function from the `rTANDEM`. This
might take some time depending on the number of files to be searched
and the search parameters. The code chunk above takes about 3 minutes
using 2 cores (**cpu = 2** above) on a modern laptop.

You can load the pre-computed quality control directory and result
data that a shipped with `proteoQC` as shown below:

```{r}
zpqc <- system.file("extdata/qc.zip", package = "proteoQC")
unzip(zpqc)
qcres <- loadmsQCres("./qc")
```


```{r}
print(qcres)
```

## Generating the QC report

The final quality report can be generated with the
**reportHTML**, passing the **qcres** object produced by
the **msQCpipe** function above or the directory storing the
QC data, as defined as parameter to the **msQCpipe**.


```{r message = FALSE}
html <- reportHTML(qcres)
```

or

```{r message = FALSE}
html <- reportHTML("./qc")
```


```{r eval=FALSE, echo=FALSE}
## Remove these files as they are really big
## but this breaks reportHTML(qcres), though
unlink("./qc/database/target_decoy.fasta")
unlink("./qc/result/*_xtandem.xml")
unlink("../inst/extdata/qc.zip")
zip("../inst/extdata/qc.zip", "./qc")
```

The report can then be opened by opening the
**qc/qc_report.html** file in a web browser or directly with
**browseURL(html)**.

# The QC report

The dynamic html report is composed of 3 sections: an introduction, a
methods and data section and a result part. The former are purely
descriptive and summarise the design matrix and analysis parameters,
as passed to **msQCpipe**.

The respective sections and sub-sections can be expanded and collapsed
and each figure in the report can be zoomed in. While the dynamic html
report is most useful for interactive inspection, it is also possible
to print the complete report for archiving.

The results section provides tables and graphics that summarise 

* Summaries of identification results for individual files as well
  as technical and biological replicates at the protein, peptide and
  spectrum levels.
* Summary overview charts that describe number of missed
  cleavages, peptide charge distributions, peptide length, precursor
  and fragment ion mass deviations, number of unique spectra/peptides
  per proteins and protein mass distributions for each sample.
* A contamination summary table generated using the common
  Repository of Adventitious Proteins (\textit{cRAP}).
* Reproducibility summaries that compare fractions, replicates and
  samples, representing total number of spectra, number of identified
  spectra, number of peptides and proteins and overlap of peptides and
  proteins across replicates.
* Summary histograms of mass accuracies for fragment and precursor
  ions.
* A summary of the separation efficiency showing the effect of
  accumulating fractions for all samples.
* A summary of identification-independent QC metrics.
 


# Some useful functions

## Protein inference

Protein inference from peptide identifications in shotgun proteomics is a very 
important task. We provide a function **proteinGroup** for this purpose.
This function is based on the method used in our another package 
`sapFinder` [@wen2014sapfinder]. You can use the function as below:

```{r fig.width=6,fig.height=5}
pep.zip <- system.file("extdata/pep.zip", package = "proteoQC")
unzip(pep.zip)
proteinGroup(file = "pep.txt", outfile = "pg.txt")
```



## Isobaric tagging reagent labeling efficiency

The labeling efficiency of the isobaric tag reagents to peptides, such as iTRAQ
and TMT, is a very important experiment quality metrics. We provide a function
**labelRatio** to calculate this metrics. You can use the function 
as below:

```{r}
mgf.zip <- system.file("extdata/mgf.zip", package = "proteoQC")
unzip(mgf.zip)
a <- labelRatio("test.mgf")
```


# Session information

All software and respective versions used to produce this document are listed below.

```{r echo=FALSE}
sessionInfo()
```

# References


[^1]: In the interest of time, the
  files are not downloaded when this vignette is compiled and the
  quality metrics are pre-computed (see details below). These
  following code chunks can nevertheless be executed to reproduce the
  complete pipeline.} 
